Exploratory data analysis of red wine by Charles Brands
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#Loading the needed libraries

library(gridExtra)
library(ggplot2)
library(ggthemes)
library(pander)
library(Hmisc)
#library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityReds.csv")
# Column X is basically the row number, remove it
wine$X <- NULL
head(wine)
```

```{r echo=FALSE, Load_the_Data}
str(wine)

# Lets look at quality
summary(wine$quality)
table(wine$quality)
```

In this analysis we look into the effect of different chemical properties on the wine quality. This dataset has 12 variabeles, the quality and 11 chemical properties of the wine (fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, and alcohol).
This dataset has 1599 observations.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
theme_set(theme_economist())
qplot(wine$quality, bins = 30)
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$quality)
```

The quality of the wines run from 3 to 8 with most of the wines are either a 5 or a 6. The quality is a discrete value. The mean is 5.6.

```{r echo=FALSE, Univariate_Plots}

geometry <- geom_histogram(bins = 30, color = 'black', fill = '#5760AB')

uni_plot <- function(plotvar, plot_label) {
  grid.arrange(
    ggplot(data = wine, aes(x=plotvar)) + geometry + labs(x = plot_label), 
    ggplot(wine, aes( x = 1, y = plotvar ) ) + 
      geom_jitter(alpha = 0.6 ) +
      geom_boxplot(alpha = 0.2, color = 'orange', size = 1) + 
      stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 8, 
               size = 4) +
      labs(y = plot_label)
  ,ncol = 2)
}

uni_plot(plotvar=wine$fixed.acidity, plot_label="Fixed acidity (g/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$fixed.acidity)
```

The fixed acidity is a right tailed distribution. The median is at 7.9 g/dm^3. Due to the outliers on the right the mean is pulled to 8.3 g/dm^3.

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$volatile.acidity, plot_label="Volatile acidity (g/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$volatile.acidity)
```

The volatile acidity is a normal distribution with a few outliers to the right. The mean and median are pretty close together at 0.5200 and 0.5278 g/dm^3 respectively.

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$citric.acid, plot_label="Citric acid (g/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$citric.acid)
```

The citric acid distribution is weird. It looks like two right skewed distribution on top of oneother.

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$residual.sugar, plot_label="Residual sugar (g/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$residual.sugar)
```

The residual sugar distribution has a sharp peak at around 2.2 g/dm^3. Although the tail is long the peak is so far above the rest that the mean is only pulled a little bit to the right of the median. The boxplot is also rather flat as the amount of wines with a residual sugar around 2.2 is huge compared to the rest.

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$chlorides, plot_label="Chlorides (g/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$chlorides)
```

The chlorides distribution is also sharply peaked. This time the peak is around 0.079 g/dm^3. 

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$free.sulfur.dioxide, plot_label="Free sulfur dioxide (mg/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$free.sulfur.dioxide)
```

The free sulfur distribution has a tail on the right. The median and mean are 14.00 and 15.87 mg/dm^3 respectively.

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$total.sulfur.dioxide, plot_label="Total sulfur dioxide (mg/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$total.sulfur.dioxide)
```

The total sulur distribution is a right tailed distribution with some outlier very far to the right. The median and the mean are 38.00 and 46.47 mg/dm^3 respectively with a max at 289 mg/dm^3!

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$density, plot_label="Density (g/cm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$density)
```

The density is a normal distributin with a mean of 0.9967 g/cm^3.

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$pH, plot_label="pH")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$pH)
```

The ph distribution is a normal distribution with a mean at 3.311

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$sulphates, plot_label="Sulphates (g/dm^3)")
```
```{r echo=FALSE, Univariate_Plots}
summary(wine$sulphates)
```

The sulfates distribution is a right skewed distributionwith some outliers far to the right. 

```{r echo=FALSE, Univariate_Plots}
uni_plot(plotvar=wine$alcohol, plot_label="Alcohol (%)")
```

```{r echo=FALSE, Univariate_Plots}
summary(wine$alcohol)
```

The alcohol percentage distribution is right skewd with a median of 10.20% and a mean of 10.42%.

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}
cor_test_with_quality <- function(x) {
  return(cor.test(x, as.numeric(wine$quality))$estimate)
}

corTable <- c(
  cor_test_with_quality(wine$fixed.acidity),
  cor_test_with_quality(wine$volatile.acidity),
  cor_test_with_quality(wine$citric.acid),
  cor_test_with_quality(wine$residual.sugar),
  cor_test_with_quality(wine$chlorides),
  cor_test_with_quality(wine$free.sulfur.dioxide),
  cor_test_with_quality(wine$total.sulfur.dioxide),
  cor_test_with_quality(wine$density),
  cor_test_with_quality(wine$pH),
  cor_test_with_quality(wine$sulphates),
  cor_test_with_quality(wine$alcohol))
names(corTable) <- c('fixed.acidity', 'volatile.acidity', 'citric.acid',
                         'residual.sugar', 'chlorides', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'sulphates', 'alcohol')
corTable
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality <- function(xplot, label) {
  grid.arrange(
    ggplot(aes(x=factor(wine$quality), y=xplot), data = wine) +
      geom_point() + 
      labs(x = "quality", y = label) + 
      geom_jitter( alpha = .3) + 
      geom_boxplot(fill="pink", color="red", alpha=0.3)  +
      coord_flip() + 
      stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 8, 
               size = 4),
    ggplot(aes(x=factor(wine$quality), y=log10(xplot)), data = wine) +
      geom_point() +
      labs(x = "quality", y = paste("log10(", label, ")")) +
      geom_jitter( alpha = .3) + 
      geom_boxplot(fill="pink", color="red", alpha=0.3)  +
      coord_flip() + 
      stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 8, 
               size = 4)
  ,ncol = 2)
}


```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$fixed.acidity, "Fixed acidity (g/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$volatile.acidity, "Volatile acidity (g/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$citric.acid, "Citric acid (g/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$residual.sugar, "Residual sugar (g/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$chlorides, "Chlorides (g/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$free.sulfur.dioxide, "Free sulfur dioxide (mg/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$total.sulfur.dioxide, "Total sulfur dioxide (mg/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$density, "Density (g/cm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$pH, "pH")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$sulphates, "Sulphates (g/dm^3)")
```

```{r echo=FALSE, Bivariate_Plots}
biv_scat_plot_quality(wine$alcohol, "Alcohol (%)")
```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
multi_plot <- function(x, y, z, alpha, size, xlabel, ylabel, zlabel) {
  breaks = c("3", "4", "5", "6", "7", "8")
  labels = c("3", "4", "5", "6", "7", "8")
  color_values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")
  ggplot(data = wine,aes(x = x, y = y, alpha = alpha)) +
    geom_point(alpha = alpha, size = size, aes(color = factor(z))) +
    xlab(xlabel) +
  ylab(ylabel) + 
  scale_color_manual(breaks = breaks,
                      labels = labels,
                      values = color_values) +
  scale_fill_manual(breaks = breaks,
                      labels = labels,
                      values = color_values)
}
```

```{r echo=FALSE, Multivariate_Plots}
multi_plot(x=wine$alcohol,y=wine$density, z=wine$quality, 
           alpha = 0.8, size=0.8,
           xlabel="Alcohol (%)", ylabel="Density (g/cm^3)",
           zlabel="Quality level")
```

```{r echo=FALSE, Multivariate_Plots}
multi_qplot <- function(x, y, z, alpha=0.4) {
  qplot(x, y, data=wine, color=z) +
    geom_point(position=position_jitter(w = 0.025, h = 0.025), alpha=alpha) +
    guides(alpha=FALSE)
}
multi_qplot(wine$density, wine$volatile.acidity, wine$quality) +
  xlab("Density (g/cm^3)") +
  ylab("Volatile acidity (g/cm^3)") +
  labs(color="Quality level")
```

```{r echo=FALSE, Multivariate_Plots}

```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!